Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada

### 2024-04-19 paper/paper.qmd
update model and paper

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
library(tinytex)
```

@@

@sec-result presents the core findings from our analysis of US birth rates, specifically focusing on the changes observed over time, across different age groups of young women, and by state.

In conclusion, our findings provide a complete perspective of dropping birth rates in the United States, demonstrating both temporal and demographic tendencies. The upcoming @sec-discussion will look at the various variables and biases that may be driving these trends.

@@

The findings of this study offer a layered perspective on the decline in birth rates among young people in the United States, particularly in the context of the Great Recession. Our analysis provides an understanding of the multifaceted influences on demographic trends.

@@

In our replication of @citeoriginal, we reaffirmed the central finding that the Great Recession has had a lasting impact on birth rates among young people in the United States. Our analysis of Figures 1 and 2 showed a pronounced decline in birth rates across all age groups, with the steepest decrease observed among those aged 15-29. This trend extended beyond the economic recovery, suggesting that the Great Recession may have fundamentally altered the family planning trajectory of a generation.

## Economic Impact Insights

The dramatic drop in birth rates among young Americans during and after the Great Recession demonstrates the profound impact of economic hardship on reproductive decisions. This decline in birth rates is not simply coincidental with the date of the recession, but rather indicates a deeper, more systemic influence of economic insecurity on personal life choices, notably the decision to establish or expand a family. The persistence of this pattern, even in the years after economic recovery, suggests that the effects of the recession went beyond immediate financial hardship, influencing long-term views of financial security and stability. This is further evidenced by the lack of rebound in birth rates post-recession, which might have been anticipated if the decline were solely due to immediate economic pressures.

The economic model of fertility provides a framework for understanding these patterns by taking into account the cost of pregnancy and raising as well as the opportunity costs associated with parental time and resources. The recession likely increased these expenditures and opportunity costs, making the decision to have children more difficult. Unemployment and job uncertainty may have caused a reevaluation of the feasibility of having children, resulting in the postponement or avoidance of childbirth. Furthermore, the economic downturn may have shifted young adults' expectations and desires for financial security, impacting their family planning decisions.

## Societal and Technological Influences

Apart from economic factors, the decline in birth rates among young people also reflects broader societal and technological shifts. The period following the Great Recession coincided with significant changes in social norms, increased educational and career opportunities for women, and advancements in reproductive technology. These factors collectively have empowered individuals, especially women, to make more independent decisions regarding their reproductive health and family planning.

As women's educational attainment and labor force involvement have increased, so has the opportunity cost of childbirth, potentially contributing to the drop in birth rates. Technological developments have given women greater control over their fertility, allowing for more intentional planning around childbirth. Social changes, such as delayed marriage and a growing acceptance of childlessness or smaller family sizes as realistic lifestyle options, further compound these tendencies.

The interaction of these economic, social, and technical elements has altered the landscape of family planning for young Americans. While the economic model of fertility provides a framework for understanding these developments, including sociological and technical effects provides a more complete view of the dynamics affecting current birth rate patterns.

## Weaknesses and Future Research Directions

One limitation of our study is the potential for unobserved variables that could affect birth rates, such as cultural shifts and changes in social norms, which were not fully captured in the data. Additionally, the original study did not account for the influence of the Affordable Care Act and its potential impact on family planning decisions, an area our study also does not explore.

Future research should aim to disentangle these complex relationships further, perhaps through longitudinal studies or by incorporating more detailed data on individual socioeconomic status. Understanding these dynamics is crucial for developing policies that support young people in their family planning decisions during and after economic downturns.

>>>>

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
library(tinytex)
library(modelsummary)
```

@@

@sec-result explores the relationship between education and average hourly wages in Canada. Utilizing a dataset that captures 20 years of wage data across varying educational levels, we apply a linear modeling approach to uncover patterns and draw conclusions about the economic value of educational advancement. Below, we present the results of our model and its implications.

```{r}
#| label: tbl-summary-model
#| tbl-cap: Model Summary
#| echo: false
#| warning: false

first_model <- readRDS(file = here::here("models/first_model.rds"))

modelsummary(
  list(
    "Linear Model" = first_model
  )
)
```

The results obtained from our linear model, which predicts the average hourly wage rate based on the numeric representation of education levels, show a clear and positive relationship between educational attainment and wage rate. @tbl-summary-model indicates that the intercept is estimated at 16.143, suggesting that when the education level is at the baseline (0 - not accounted for in the numeric scale), the average hourly wage rate would be approximately 16.14. Importantly, the coefficient for Education_numeric is positive (1.672), supporting the hypothesis that higher educational levels correlate with higher wage rates. The model used 180 observations for the analysis, with an R-squared value of 0.610, which means that approximately 61% of the variance in the average hourly wage rate is explained by the model. The adjusted R-squared, which accounts for the number of predictors in the model, is very close to the R-squared, indicating a good fit without overfitting. Notably, the RMSE (Root Mean Square Error) of 3.43 reflects the standard deviation of the prediction errors, which measures how spread out these errors are around the true regression line.

```{r fig.width=7, fig.height=4}
#| label: fig-visualize-model
#| fig-cap: Model Visualization
#| echo: false
#| warning: false

analysis_data <- read.csv(file = here::here("data/analysis_data/analysis_data.csv"))

plot1 <- analysis_data |>
  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate, colour = as.factor(Year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  scale_colour_viridis_d() +  # Colorblind-friendly palette
  labs(
    x = "Education Level",  # Rename x-axis
    y = "AVG Hourly Wage Rate",  # Rename y-axis
    colour = "Year"  # Rename legend title
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_text(size = 12),  # Adjust x-axis title size if needed
    axis.title.y = element_text(size = 12),  # Adjust y-axis title size if needed
    legend.position = "right",
    legend.title = element_text(hjust = 0.5)
  ) +
  guides(colour = guide_legend(title = "Year", ncol = 2))
# Display the plot
print(plot1)
```
@fig-visualize-model illustrates the fitted linear regression line across the spread of average hourly wage rates for different education levels over the years 2000 to 2019. Each point in the scatter plot corresponds to the average hourly wage rate for a given education level, differentiated by color to represent each year. The dashed line across the plot signifies the best-fit linear regression line, which inclines upward as the education level increases. The diversity of colors along the regression line also reveals the variation in wage rates across different years, although the increasing trend is consistent throughout.

In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate. This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities. The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion. In the next @sec-discussion, we will delve into the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.

@@

In this paper, we investigated the relationship between educational attainment and average hourly wage rates from 2000 to 2019. By deploying a linear Bayesian model, we have been able to highlight the correlation between higher educational attainment and increased wage rates, aligning with our original hypothesis.

@@

Our findings reinforce the perception of education as an investment yielding tangible financial returns. The analysis showcases a clear, positive trajectory in wage rates corresponding with ascending educational levels. This outcome underlines the vital role of education in driving economic growth, suggesting that educational enhancements could be pivotal in strategies aimed at combating poverty and fostering economic resilience.

Further examination reveals the potential of education as a tool for social mobility. The data indicates that as individuals climb the educational ladder, they unlock higher earnings, which could lead to a more equitable distribution of income. These insights could serve as a valuable asset for policymakers striving for educational equity and highlight the importance of making higher education accessible to all strata of society.

## Insights on Education and Wages

The positive correlation observed in this study goes beyond confirming a long-suspected link between education and wages; it also points to the complex interplay of various wage determinants. Despite the clear trend, the dispersion of wage data around the regression line suggests multiple underlying factors at play. This finding suggests a need for a comprehensive approach to wage enhancement, integrating elements such as vocational training, relocation incentives, and specialized educational programs.

Expanding on the above, it is evident that education serves as one of many pillars supporting an individual's earning capacity. While education remains a significant factor, the disparity between the predicted and observed wages implies other variables are also influential. Recognizing this complexity is crucial for a complete understanding of how education impacts wage rates and the formulation of holistic policies that address the multifaceted nature of the job market.

## Limitations

The primary limitation of this study is its relatively small sample size, with only 180 observations. This constraint may impact the robustness and generalizability of the findings. Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers. Such a limited scope might overlook the nuanced effects of education on wages across different life stages.

## Future Research Directions

Future research should aim to disaggregate the age groups to understand the differential impact of education on wages across a more granular age spectrum. Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more nuanced insights into the education-wage dynamic at different life stages. Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.

Incorporating a more comprehensive range of variables and exploring more complex, perhaps non-linear models, would allow for a better understanding of the multifarious factors influencing wages. Such an enriched analysis could serve as the foundation for more effective, tailored economic policies that address both the direct and indirect ways education can influence economic outcomes.
